---
title: "Lung Epithelial Clock Training"
author: "Xiaolong Guo, Andrew E. Teschendorff"
date: "`r Sys.Date()`"
package: CellAgeR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Lung Epithelial Clock Training}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Overview

This workflow shows a training pipeline for a lung epithelial
clock. The workflow first estimates epithelial fractions with EpiDISH,
then restricts the analysis to high cell type specificity score (CSS)
lung epithelial markers, prioritizes lung epithelial specific age DMCs
with Sliding Window Random Forest (SWRF), and finally trains a Lung Epithelial Clock.

1.  Estimate epithelial fractions with `EpiDISH`.
2.  Extract lung epithelial DMCs with `CSS > 0.4`.
3.  Find lung epithelial specific age DMCs with `SWRF`.
4.  Train a lung epithelial specific clock with `trainWeightedClock()`.

## Setup

```{r setup, message=FALSE, warning=FALSE}
pkg_name <- "CellAgeR"

if (!exists("SWRF", mode = "function")) {
  library(pkg_name, character.only = TRUE)
}

library(EpiDISH)
library(ggplot2)

example_dir <- system.file("extdata", package = pkg_name)
if (!nzchar(example_dir)) {
  example_dir <- "inst/extdata"
}
```

## Load Data

The training file contains two objects: `lungMat`, a methylation matrix
with CpGs in rows and samples in columns, and `ageVector`, a numeric
vector of chronological ages aligned to the columns of `lungMat`. The
workflow also loads `cssSignatures` and extracts lung epithelial CpGs
with `CSS > 0.4` as the initial candidate feature set.

```{r load-data}
load(file.path(example_dir, "lungGSE175458.rda"))
data("cssSignatures", package = pkg_name)
lung_dmcs <- names(cssSignatures$lungEpi[cssSignatures$lungEpi > 0.4])
```

```{r load-data-prep, include=FALSE}
lungMat <- lungMat / 10000
```

## Estimate Cell-Type Fractions with EpiDISH

`centEpiFibIC.m` is a reference for estimating epithelial, fibroblast,
and immune cell fractions.

```{r epidish-fit, message=FALSE}
cell_fractions <- as.data.frame(epidish(
  beta.m = lungMat,
  ref.m = centEpiFibIC.m,
  method = "RPC"
)$estF)
fraction_df <- stack(cell_fractions)
colnames(fraction_df) <- c("Fraction", "CellType")

ggplot(fraction_df, aes(x = CellType, y = Fraction, fill = CellType)) +
  geom_boxplot(alpha = 0.7, outlier.size = 0.3) +
  theme_bw() +
  labs(
    x = NULL,
    y = "Estimated fraction"
  ) +
  theme(legend.position = "none")
```

## Run SWRF to find Lung Epithelial specific Age DMCs

```{r swrf-run}
swrf_res <- SWRF(
  targetCellType = "Epi",
  candidateFeatures = lung_dmcs,
  dataMatrix = lungMat,
  cellFractions = cell_fractions,
  responseVector = ageVector
)
candidate_data <- subset(swrf_res, Correlation > 0.1)
head(candidate_data)
```

## Convert SCCs into Penalty Factors

```{r penalties}
penalties <- 1 / abs(candidate_data$Correlation)
penalties <- penalties / mean(penalties)
names(penalties) <- candidate_data$Feature
```

## Train a Lung Epithelial Specific Clock

```{r train-clock}
clock_fit <- trainWeightedClock(
  dataMatrix = lungMat,
  responseVector = ageVector,
  candidateFeatures = candidate_data$Feature,
  featureWeights = penalties
)
glmnet_fit <- clock_fit$cvModel
head(clock_fit$clockCoefficients)
```

```{r lambda-plot}
plot(glmnet_fit)
```

```{r validation}
lambda_idx <- which.min(abs(glmnet_fit$lambda - glmnet_fit$lambda.min))
y_cv_pred <- glmnet_fit$fit.preval[, lambda_idx]

validation_df <- data.frame(
  Actual_Age = ageVector,
  Predicted_Age = y_cv_pred
)

ggplot(validation_df, aes(x = Actual_Age, y = Predicted_Age)) +
  geom_point(alpha = 0.7, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "firebrick") +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  theme_bw() +
  labs(
    title = "lung (n=201, GSE175458, Training set)",
    x = "Actual age",
    y = "Cross-validated predicted age"
  )
```

## Extract Clock Coefficients

The trained clock coefficients are returned directly by
`trainWeightedClock()` and can be extracted from the fitted result for
inspection or downstream prediction.

```{r extract-coefficients}
clock_coefficients <- clock_fit$clockCoefficients
head(clock_coefficients)
```

## Session Info

```{r sessionInfo}
sessionInfo()
```
