---
title: "Introduction to CellAgeR: Enhancer-Enriched Cell-Type Specific DNA Methylation Clocks"
author: "Xiaolong Guo, Andrew E. Teschendorff"
date: "`r Sys.Date()`"
package: CellAgeR
output:
  BiocStyle::html_document:
    toc: true
    toc_float: true
    number_sections: true
vignette: >
  %\VignetteIndexEntry{Introduction to CellAgeR: Enhancer-Enriched Cell-Type Specific DNA Methylation Clocks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center"
)
```

## Introduction

The `CellAgeR` package provides a suite of enhancer-enriched, cell-type specific DNAm clocks. Biological aging often manifests differently across distinct cell types, yet standard bulk-tissue clocks can be confounded by changes in cellular heterogeneity.

By focusing on DNA methylation patterns at enhancer-enriched cell-type specific regions, `CellAgeR` enables quantification of biological age for: 
- **Brain**: Neurons and Oligodendrocytes (and sorted versions)
- **Epithelial cells**: Skin Keratinocytes, Colon, Lung, and Saliva
- **Liver**: Hepatocytes
- **Immune cells**: CD4+ T cells, Monocytes, Colon, Lung, and Saliva Immune cells
- **Fibroblasts**: Lung and Colon Fibroblasts

This vignette demonstrates how to use the core functions to predict epigenetic age and perform downstream analysis using example datasets.

## Setup

```{r setup, message=FALSE, warning=FALSE}
pkg_name <- "CellAgeR"

if (!exists("getClockCoefficients", mode = "function")) {
  library(pkg_name, character.only = TRUE)
}

library(ggplot2)
library(patchwork)
library(ggpubr)

example_dir <- system.file("extdata", package = pkg_name)
if (!nzchar(example_dir)) {
  example_dir <- "inst/extdata"
}
```

## Data Preparation

`CellAgeR` requires a DNA methylation beta matrix with CpGs in rows and samples in columns, plus a clock model. The package includes pre-trained models that can be retrieved with `getClockCoefficients()`.

```{r data_check}
clock_coefficients <- getClockCoefficients()
clock_env <- list2env(clock_coefficients, parent = emptyenv())

# Available clocks in the package
ls(clock_env)
```

The packaged example datasets follow the same structure used in the tests:

-   `betaMat`: methylation beta matrix
-   `phenoDf`: sample-level phenotype table

## Quick Start: Predicting Epigenetic Age

In this example, we use the `Walker et al.` dataset which contains sorted Neurons (`NeuNPos`), Oligodendrocytes (`Sox10Pos`), and bulk Prefrontal Cortex (PFC) samples (`Total`).

We will predict age using both the `NeuronClock` and the `OligoClock`, following the same workflow used in the validated test script.

```{r quick_start, message=FALSE}
# Load example data
load(file.path(example_dir, "Walker.Rdata"))

clocks <- list(
  Neuron = clock_coefficients$NeuronClock,
  Oligo = clock_coefficients$OligoClock
)

cell_types <- c(
  "NeuNPos" = "Neuron",
  "Sox10Pos" = "Oligo",
  "Total" = "PFC"
)

plot_list <- list()

for (clock_name in names(clocks)) {
  for (ct_key in names(cell_types)) {
    ct_label <- cell_types[[ct_key]]
    idx <- which(phenoDf$CellType == ct_key)
    p_sub <- phenoDf[idx, ]
    b_sub <- betaMat[, idx]

    res <- predictCTSAge(b_sub, clocks[[clock_name]], age = p_sub$Age)

    pred_df <- res$predictions
    real_age <- p_sub$Age
    pred_age <- pred_df$DNAmAge
    
    valid <- !is.na(pred_age) & !is.na(real_age)
    real_age <- real_age[valid]
    pred_age <- pred_age[valid]
    
    if (length(real_age) > 2) {
      pcc <- cor(real_age, pred_age, method = "pearson")
      medae_val <- median(abs(real_age - pred_age))
      slope <- coef(lm(pred_age ~ real_age))[2]

      df_plot <- data.frame(Age = real_age, Predicted = pred_age)
      range_val <- range(c(real_age, pred_age), na.rm = TRUE)
      annot_str <- sprintf(
        "PCC = %.2f\nMedAE = %.2f\nSlope = %.2f",
        pcc,
        medae_val,
        slope
      )

      p <- ggplot(df_plot, aes(x = Age, y = Predicted)) +
        geom_point(alpha = 0.6, color = "steelblue", size = 0.5) +
        geom_abline(intercept = 0, slope = 1, color = "darkgrey", linetype = "dashed") +
        geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 0.5) +
        annotate(
          "text",
          x = -Inf,
          y = Inf,
          label = annot_str,
          hjust = -0.1,
          vjust = 1.2,
          size = 2.47
        ) +
        labs(
          title = paste0(clock_name, "Clock -> ", ct_label),
          x = "Chronological Age",
          y = "Predicted Age"
        ) +
        coord_fixed(xlim = range_val, ylim = range_val) +
        theme_bw() +
        theme(
          plot.title = element_text(size = 7, face = "bold", hjust = 0.5),
          axis.title = element_text(size = 7),
          axis.text = element_text(size = 7)
        )

      plot_list[[paste(clock_name, ct_key, sep = "_")]] <- p
    }
  }
}

wrap_plots(plot_list, ncol = 3, nrow = 2)
```

## Downstream Analysis: Age Acceleration in Huntington's Disease

A key application of epigenetic clocks is calculating age acceleration, defined here as the residual from regressing DNAmAge on chronological age. Using the `GSE284764` dataset, we test whether striatal neurons from Huntington's disease cases show elevated age acceleration relative to controls.

```{r downstream, message=FALSE, warning=FALSE}
# Load HD dataset
load(file.path(example_dir, "HDStriatum.Rdata"))

# Predict with NeuronClock
res_hd <- predictCTSAge(
  betaMatrix = betaMat,
  clock = clock_coefficients$NeuronClock,
  age = phenoDf$Age
)

# Prepare dataframe for analysis
plot_df <- data.frame(
  Group = phenoDf$GroupRaw,
  AgeAccel = res_hd$ageAcceleration$AgeAcceleration
)

# Visualization & statistical annotation
ggplot(plot_df, aes(x = Group, y = AgeAccel, fill = Group)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.5, size = 0.3) +
  geom_jitter(width = 0.2, alpha = 0.7, size = 0.5) +
  stat_compare_means(
    method = "wilcox.test",
    method.args = list(alternative = "less"),
    label.x.npc = "center",
    label.y.npc = "top",
    size = 2.47,
    aes(label = paste0("P = ", ..p.format..))
  ) +
  theme_bw() +
  labs(title = "Neuron clock -> Striatum",
       y = "Residuals Age Acceleration",
       x = NULL) +
  theme(legend.position = "none",
        plot.title = element_text(hjust = 0.5, face = "bold", size = 7),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_fill_manual(values = c("Control" = "steelblue", "Huntington's disease" = "firebrick"))
```

## See Also

For a step-by-step training workflow that uses `EpiDISH`, `cssSignatures`,
`SWRF()`, and `trainWeightedClock()` to build a lung epithelial clock, see
[Lung Epithelial Clock Training](lung-workflow.html).

## Session Info

```{r sessionInfo}
sessionInfo()
```
